On Optimality of Finite Element Integration

### 2015-10-30 TOMS_Paper/assembly.tex
Improve cost function and trade opt sections

At this point, two main issues are to be addressed: 
\begin{enumerate}
\item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation, which is legal only under a specific factorization of the expression, could either increase or decrease the operation count with respect to sharing elimination.
\item \textit{Finding the global optimum.} Consider a form comprising two monomials $m1$ and $m2$. Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g. some basis functions). If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost. We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
\end{enumerate}

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss of pre-evaluation over sharing elimination. In particular, we have that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$. $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying to sharing elimination and pre-evaluation, respectively. We return to the issue of constructing $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}. $\theta$ is used by the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.

@@

// Find optimum such that memory constraints are honoured
B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
for each b $\in$ B:
  for each (b$_S$, b$_P$) $\in$ b:
    if memory_required(b$_P$) $\geq$ $T_H$:
      continue
    op_count[b] = $\theta_{se}$(S $\cup$ b$_S$) + $\theta_{pre}$(b$_P$)
(b$_S$, b$_P$) = min(op_count)

@@

The algorithm starts with splitting the monomials into two disjoint sets: the monomials that surely do not take advantage of pre-evaluation ($S$) and those that instead potentially benefit from it ($P$). Some subsets of monomials in $P$: 1) might share terms with $S$, in which case sharing elimination could globally outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}. The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of monomials for which pre-evaluation is globally optimal and Constraint~\ref{const:TH} is honoured. Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible. This aspect is elaborated in Section~\ref{sec:op_count}. In addition, observe that because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary. Therefore, sharing elimination is transparently applied to all monomials (i.e., to all sub-expressions, even those produced by pre-evaluation). The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.

@@

In the following section, we tie up the remaining loose end, the construction of the cost function $\theta$.

\subsection{The cost function $\theta$}
\label{sec:op_count}
We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and with pre-evaluation. Since $\theta$ is expected to be used by a compiler, requirements are simplicity and velocity. In the following, we explain how to derive these two values.

The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the transformations and count the number of operations. If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} tends to have negligible cost), the code generation time would become unacceptable if we were to apply pre-evaluation (Algorithm~\ref{algo:pre-evaluation}, symbolic execution of $L_i$) to each possible bipartition. We then need an analytic way of determining $\theta_{pre}$.

The first step consists of estimating the \textit{increase factor}, $\iota$. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}. One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities. By product expansion, we can observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor 3 (the number of basis functions for $f$). $\iota$ captures this increase in non-hoistable operations.

@@

With a single coefficient, as we just saw, $\iota$ directly descends from the cost of product expansion. In general, however, predicting $\iota$ is less straightforward. For example, consider the case in which a monomial has multiple coefficients over the same function space. Product expansion would lead to identical sub-expressions, and $\iota$ must reflect this. To evaluate $\iota$, we then calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$) and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$).

If $\iota \geq I$ we can immediately say that pre-evaluation will not be profitable. This is a necessary condition telling us that if we add to the innermost loop more operations than we actually save from eliminating $L_i$, then for sure $\theta_{se} < \theta_{pre}$. This observation can speed up the analysis cost.

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{pre}$. In particular, we need to calculate the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators used in the form. For example, in the case of the bilinear form originating from a standard bi-dimensional Poisson equation, the reader could verify that after a suitable factorization we would have $\rho = 3$. To determine $\rho$, we simply apply Algorithm~\ref{algo:pre-evaluation} until the evaluation of the reduction (obviously, excluded).

\subsection{Trading optimality for analysis cost}
Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing the constraints on optimality.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts render certain factorizations sub-optimal. To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level. Since this situation is particularly rare, we can relieve the algorithm by avoiding these checks.
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible ways of scheduling monomials candidate to pre-evaluation. In extremely complex forms, this could increase the analysis cost. Despite we have never encountered such a situation, a possibility is to disable these checks and to apply the transformation to all candidates. In the worst case scenario, this leads to a local optimum.
\item \textit{Construction of $\theta$.} Instead of determining $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by manipulating the expression tree, we can approximate these values exploiting the key property of outer product loop nests, linearity. The number of unique terms in a multilinear loop nest can be used to calculate an upper bound on the cost of the innermost loop, where the bulk of the computation naturally tends to be performed. For example, in Figure~\ref{code:multi_loopnest_a} the fact that only a single term depends on $L_j$, $b_j$, suggests that in the innermost loop no more than one multiplication will be performed once sharing is eliminated. Therefore, an approximation for $\theta_{se}$ is, for example, $\theta_{se} = n_{iterations} n_{unique}$, where $n_{iterations}$ is the total number of iterations, and $n_{unique}$ is the smallest number of unique terms along one the linear loops.
\end{itemize}

In our code generation system, described in the next section, we make use of all of these heuristics.

\section{Code Generation}
\label{sec:codegen}
The transformation algorithm described in Section~\ref{sec:optimal-synthesis} has been implemented in COFFEE, the optimizer of finite element integration routines used in Firedrake. In this section, we describe several aspects related to code generation.

>>>>

At this point, two main issues are to be addressed: 
\begin{enumerate}
\item \textit{Coordinating the application of pre-evaluation and sharing elimination.} Recall from Section~\ref{sec:pre-evaluation} that pre-evaluation could either increase or decrease the operation count with respect to sharing elimination.
\item \textit{Finding the global optimum.} Consider a form comprising two monomials $m1$ and $m2$. Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g. some basis functions). If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost. We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- results in the highest operation count reduction when considering the whole set of monomials (i.e., the expression as a whole).
\end{enumerate}

Let $\theta : M \rightarrow \mathbb{Z}$ be a cost function that, given a monomial $m \in M$, returns the gain/loss achieved by pre-evaluation over sharing elimination. In particular, we define $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, where $\theta_{se}$ and $\theta_{pre}$ represent the operation counts resulting from applying sharing elimination and pre-evaluation, respectively. Thus pre-evaluation is profitable for $m$ if and only if $\theta(m) > 0$. We return to the issue of deriving $\theta_{se}$ and $\theta_{pre}$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can focus on the transformation algorithm, whose pseudo-code is provided in Figure~\ref{code:intuition}.

@@

// Find optimum such that memory constraints are honored
B = $\lbrace$b | b is a bipartite graph from complete graph G = (P, E)$\rbrace$
for each b $\in$ B:
  for each (b$_S$, b$_P$) $\in$ b:
    if memory_required(b$_P$) $\geq$ $T_H$:
      continue
    op_count[b] = $\theta_{se}$(S $\cup$ b$_S$) + $\theta_{pre}$(b$_P$)
(b$_S$, b$_P$) = min(op_count)

@@

The algorithm starts with splitting the monomials into two disjoint sets: the monomials that surely do not take advantage of pre-evaluation ($S$) and those that instead potentially benefit from it ($P$). Some subsets of monomials in $P$: 1) might share terms with $S$, in which case sharing elimination could globally outperform pre-evaluation; 2) might break Constraint~\ref{const:TH}. The algorithm then finds the optimal bipartition $b = (b_S, b_p)$ of $P$ such that $b_P$ represents the subset of monomials for which pre-evaluation is globally optimal and Constraint~\ref{const:TH} is honored. Since the number of monomials in a form tends to be very small (a few units in most complex forms), evaluating all possible bipartitions is not a significant issue as long as the cost of calculating $\theta_{se}$ and $\theta_{pre}$ is negligible. This aspect is elaborated in Section~\ref{sec:op_count}. In addition, observe that because of reuse of basis functions, pre-evaluation may result in identical tables, which will be mapped to the same temporary. Therefore, sharing elimination is transparently applied to all monomials (i.e., to all sub-expressions, even those produced by pre-evaluation). The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming as input the loop nest in Figure~\ref{code:loopnest}.

@@

In the following section, we tie up the remaining loose end: the construction of the cost function $\theta$.

\subsection{The cost function $\theta$}
\label{sec:op_count}
We recall that $\theta(m) = \theta_{se}(m) - \theta_{pre}(m)$, with $\theta_{se}$ and $\theta_{pre}$ representing the operation counts after applying sharing elimination and pre-evaluation. Since $\theta$ is expected to be used by a compiler, requirements are simplicity and velocity. In the following, we explain how to derive these two values.

The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations. If, on one hand, this is definitely possible for $\theta_{se}$ (performing Algorithm~\ref{algo:opt} tends to have negligible cost), the overhead becomes unacceptable when applying pre-evaluation to all possible bipartitions (Algorithm~\ref{algo:pre-evaluation}, due to the symbolic evaluation of $L_i$). We then seek an analytic way of determining $\theta_{pre}$.

The first step consists of estimating the \textit{increase factor}, $\iota$. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}. One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities. By performing product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.

@@

Despite determining $\iota$ is straightforward in simple scenarios (e.g., when the form has a single coefficient, as we just saw), in general we must account for repeated sub-expressions that, once pre-evaluated, would result in identical tables. This is the case, for example, of a monomial having more than one coefficient in the same function space. One can verify this by taking the example in Figure~\ref{code:increase_factor} and adding a second coefficient $g$ over the same basis functions $B$. We then use combinatorics to evaluate $\iota$. We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product (in the example, there is only $f$, so $k=1$); (ii) $n$ is the number of unique basis functions involved in the expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$).

If $\iota \geq I$ (the extent of the reduction loop), we already know that pre-evaluation will not be profitable. The condition $\iota < I$ is necessary to obtain a reduction in operation count. Intuitively, if this were not true, it would mean that we would be introducing more operations than we would save from pre-evaluating $L_i$. This observation can speed up the analysis cost.

With $\iota$ in place, it remains to find the total number of terms, $\rho$, that will be affected by the increase factor. Then we have $\theta_{pre} = \rho \cdot \iota$. Consider again the integration for the mass matrix in Figure~\ref{code:increase_factor}. This form is characterized by the dot product of test and trial functions (\texttt{B[j]*B[k]}), so here, trivially, $\rho = 1$. In general, $\rho$ depends on both form and discretization. As another example, take a standard Poisson equation using P1 Lagrange elements in 2D: we have $\rho = 3$ since there are in total four spatial derivatives and factorizing the expression as explained in Algorithm~\ref{algo:pre-evaluation} leads to 3 integration-dependent terms. We determine $\rho$ by executing parts of Algorithm~\ref{algo:pre-evaluation}, which obviously excludes the reduction pre-evaluation.

\subsection{Trading optimality for analysis cost}
Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing optimality.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts may render certain factorizations sub-optimal. To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level. Since such increases are, in practice, rare and negligible, the algorithm can be relieved by avoiding these checks.
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints. Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials. This matter is yet to be properly investigated. However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity. The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed). For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated. An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
\end{itemize}

These heuristics have been implemented in the code generation system described in Section~\ref{sec:codegen}.

\section{Code Generation}
\label{sec:codegen}
Sharing elimination, pre-evaluation, the transformation algorithm, and all other features presented in Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} have been implemented in COFFEE, the optimizer for finite element integration routines used in Firedrake. In this section, we discuss implementation and code generation.
